---
title: "Fertility in Skellefteå, mothers born 1820--1869"
author: "Göran Broström"
date: '2015-02-26'
output:
  html_document:
    fig_caption: yes
    toc: yes
    toc_depth: 3
  pdf_document:
    fig_caption: yes
    toc: yes
---

## Introduction, Skellefteå data.

This version gives a more restrictive data set; we require that all mothers are born in Skellefteå.

The goal is to get a data file with fertility histories of women in Skellefteå. We are only interested in women born between 1 January, 1820 and 31 december 1869, and that gave birth to at least one child. So,

```{r skel14}
    library(skel14) # Data for Skellefteå, from POPLINK.
    females <- person[with(person, KON == 2 & 
                               FODDAT > 18200000 & 
                               FODDAT < 18700000), ]
    ### New:
    oid <- observation$ID[with(observation, STARTTYP == 2)]
    females <- females[females$ID %in% oid, ]
    ## End New
    mothers <- females[females$ID %in% person$MID, ]
    n.moth <- length(unique(mothers$ID)) # 'unique' not necessary here.
```

So there are `r n.moth` mothers which we can have as a starting point. Let us now collect all kids to these mothers.
```{r kids}
infants <- person[person$MID %in% mothers$ID, ]
n.infants <- length(unique(infants$ID))
infants <- infants[order(infants$MID, infants$FODDAT), ]
##save(infants, file = "infants.rda")
```

Now, from this we make a file with *all births*, letting *id* be the id of the mother, and *ch_id* the id of the child.

```{r births}
indx <- match(infants$MID, mothers$ID)
births <- data.frame(id = infants$MID, 
                     birthdate = mothers$FODDAT[indx], 
                     day = infants$FODDAT, 
                     parity = infants$PARITET_G, 
                     ch_id = infants$ID, 
                     sex = infants$KON, 
                     ab = infants$AB, 
                     df = infants$DF, 
                     fb = infants$FB)
births  <- births[order(births$id, births$day), ] # Sorting by mother and birth order within mother
##save(births, file = "births.rda")
n.births <- length(unique(births$ch_id))
n.mothers <- length(unique(births$id))
```

Ok, we have `r n.births` births to these `r n.mothers` mothers (`r round(n.births / n.mothers, digits = 1)` births per mother on average). Let us look at what we have so far (the first rows of *births*):

```{r head}
head(births, n = 10)
```

The variables are: 

*  **id**: Mother's id.
*  **birthdate**: Mother's birthdate.
*  **day**: This child's birthdate.
*  **parity**: The birth order of the child (from source, "*PARITET_G*, "genererat bland biologiska barn till modern").
*  **ch_id**: This child's id.
*  **sex**: Sex of the child, 1 = 'boy', 2 = 'girl', 3 = 'hermaphrodite', 0 = 'unknown'.
*  **ab**: Mother's civil status at delivery: 1 = 'married', 2 = 'unmarried', 3 = 'engaged', 0 = 'unknown'.
*  **df**: Stillbirth? 0 = 'no', 1 = 'yes'.
*  **fb**: Multiple birth: 1 = 'single (or unknown)', 2 = 'twin', 3 = 'triplet'.

## Defining factors 

In order to get more informative summaries, we define some key variables as *factors*:

```{r fack}
## births$id <- as.character(births$id) # Dangerous!! (sorting...) A note to self...
##births$ch_id <- as.character(births$ch_id)
births$sex  <- factor(births$sex, levels = c(1, 2, 3), labels = c("boy", "girl", "hermaphrodite"))
births$ab <- factor(births$ab, levels = 1:3, labels = c("married", "unmarried", "engaged"))
births$stillbirth <- factor(births$df, labels = c("no", "yes"))
births$mult_birth <- factor(births$fb, levels = 1:3, labels = c("no", "twin", "triplet"))
births$df <- births$fb <- NULL
```

Now, take a look at the result:

```{r look}
summary(births)
```

## Removing mothers with incomplete data

Some mothers lack a birth date (only year is given), and we remove them from the data set. They are characterized by the fact that the fifth and sixth digits of *birthdate* are both zero. We can easily check for that utilizing the *mode* function and integer division.

```{r mode}
bda <- births$birthdate %/% 100
weq <- bda %% 100 == 0
id.weq <- unique(births$id[weq])

births <- births[!(births$id %in% id.weq), ]
n.newmr <- length(unique(births$id))
summary(births)
```

After this, we are left with `r n.newmr` mothers, that is, `r n.mothers - n.newmr` left us. There are still mothers left with known birth month but unknown day within the month (two last digits are both zero). We keep them and let them have a birthdate in the middle of the month:

```{r midmon}
births$birthdate <- births$birthdate + 15 * (births$birthdate %% 100 == 0) # TRUE --> 1, FALSE --> 0.
```

## Mother's age at child births

Now we will create a new variable, **age**, which will be mother's age at the birth of each child. It is essentially the difference between *day* and *birthdate*, but first we must fix bad values in *day*, and then translate dates to (fractions of) years.

Regarding the dates of child births, they are so essential, that we remove all mothers who have *any* bad *day* value. This includes mothers with *day = 0* and with the last two digits of *day* both equal to zero.

```{r daysgone}
id.weq <- unique(with(births, id[day %% 100 == 0]))
births <- births[!(births$id %in% id.weq), ]
n.bir <- length(unique(births$id))

```

After this operation, we are left with `r n.bir` mothers and lost `r n.newmr - n.bir`; not too bad. It remains to fix the *subtraction*:

```{r subtract}
births$age <- tillTid(births$day) - tillTid(births$birthdate)
summary(births)
##save(births, file = "births.rda")
```

## Annoyances

We have some annoyances left, the missing values on *sex*, *ab*, and *mult_births*. Regarding missing on *sex*, it is strongly connected to *stillbirth*:

```{r tablesexstill}
with(births, table(stillbirth, sex, useNA = "ifany"))
```

Let us identify those women who have births without gender, but not stillbirth:

```{r stillsex}
(mid <- births$id[births$stillbirth == "no" & is.na(births$sex)])
```

I decided to remove those `r length(mid)` mothers (and their births) from the data file. A the same time, some *cleaning* is performed: 

*  There are no *hermaphrodites*, so that category is removed,

*  Those with missing *mult_birth* are assumed to be *no* mult_birth.

*  For *ab*, we assume that *missing* is *married*, and *engaged* are regarded as *married*.

```{r herma}
births <- births[!(births$id %in% mid), ]
births$sex <- factor(births$sex)
births$mult_birth[is.na(births$mult_birth)] <- "no"
births$ab[is.na(births$ab)] <- "married"
births$ab[births$ab == "engaged"] <- "married"
births$ab <- factor(births$ab)
births <- births[order(births$id, births$age), ] ## New, does it help??
summary(births)
```

Only two disturbing facts are left now; can you spot them?

1. The number of twins is an *odd number*, it should be even, and

2. At least one mother is above 57 years of age at a birth. And one is below 12 (not completely unrealistic?).

## Multiple births

Let us first check the twin births:

```{r twinbirths}
mtw <- with(births, tapply(mult_birth == "twin", id, sum))
(xx <- table(mtw))
```

So, the vast majority of the mothers had no twins, `r xx[2]` had one, `r xx[3]` had two, `r xx[4]` three, `r xx[5]` four, and 
`r xx[6]` had  eight (she had *four* twin births!). 

Let's have a look at those having three twins:

```{r twinmothers}
(mtw3 <- mtw[mtw == 3])
```

The births for the first mother, id = `r names(mtw3)[1]`, are

```{r fmor}
births[births$id %in% names(mtw3)[1], c("id", "parity", "sex", "mult_birth", "age", "stillbirth")]
```

Obviously, she got four twins (two twin births), not three. The second line is erroneously marked as no twin birth.
Note also that the variables *parity* and *age* both show that the first two lines constitute a twin birth.
The variable *mult_birth* is thus not quite reliable.

With the aid of the function *is.inc* (short for *"is increasing?"*) we can check the reliability of the different variables for *birth order*:

```{r inc}
source("R/is.inc.R")
is.inc
xx <- with(births, tapply(parity, id, is.inc))
table(xx)
```

OK, it would be surprising to get another result. Let us check *age*:

```{r age}
xx <- with(births, tapply(age, id, is.inc))
table(xx)
```

Let us give an alternative definition of *multiple birth*: If the difference in birth dates is less than one week (approximately 0.02 years), it is a twin or triplet birth. The function *mult* helps us:

```{r mult}
source("R/mult.R")
mult
```

This is the definition, now we apply it and create a new variable *multi*, with values 0, 1, 2, 3:

```{r multi}
indx <- with(births, tapply(age, id))
multi <- unlist(with(births, tapply(age, id, mult)))
births$multi <- multi
##source("R/mult3.R")
##births <- mult3(births)
head(births[, c("id", "parity", "stillbirth", "mult_birth", "multi")], 10)
with(births, table(multi, mult_birth, useNA = "ifany"))
births$mult_birth[births$id == 823500 & births$parity == 6] <- "triplet" ## Correct a booking error
```

## Sex distribution

We now want the cumulative numbers of girls and boys (surviving birth) for each woman: 

```{r sex}
boy <- births$sex == "boy" & births$stillbirth == "no"
girl <- births$sex == "girl" & births$stillbirth == "no"
births$boys <- unlist(tapply(boy, births$id, cumsum))
births$girls <- unlist(tapply(girl, births$id, cumsum))
```

## Transform file to describe intervals

### Make one record of multiple births

Next we will transform the data frame *births* to a data frame of births and not of children. That is twin and triplet birts should be recorded as one birth, but marked with the approprate attributes. We therefore remove all rows with *multi == 1* and all rows with *multi == 2 AND mult_birth == "triplet"*:

```{r remtrtw}
weq <- with(births, (multi == 1) | ((multi == 2) & (mult_birth == "triplet")))
births <- births[!weq, ]
rownames(births) <- 1:NROW(births)
head(births, 10)
```

## Add *enter*, *exit*, *event*; "too old" mothers

Now, this is what our data look like right now. We need to turn these events into intervals between births. The variable *age* is the start of the interval and the end of the interval is the age of birth on the *next* interval (for the same woman). For her last interval we put in an exit age of 50 or the age of her disappearance if before age 50. For the time being we put in age 50. Turns out to be a small problem with this: Three women gave birth at ages above 50, the oldest was 57. This is *very* rare, especially in historical times,
so we remove the three intervals with enter *age >= 50*. They all end with *no birth*.

Since our data is ordered by woman and by age within woman, we get the exit variable by simply lagging the age variable and give it a new name, *exit*; *enter* is the actual age:

```{r entex}
births$enter <- births$age
births$exit <- c(births$age[-1], 50)
births <- rc(births) # Adds 'antrec' and 'lopnr' to 'births'.
head(births[, c("id", "age", "enter", "exit", "lopnr", "antrec")])
births$exit[births$antrec == births$lopnr] <- 50 ## NOTE row 5 above!
births$event <- with(births, antrec != lopnr)
births <- births[births$enter < 50, ]
save(births, file = "births.rda")
```

## Get mother's "last seen" age

We need to replace *exit == 50* with a real exit age; We fetch it from the *person* data frame:

```{r getperson}
lastseen <- person$UTDATMIN[match(births$id, person$ID)]
births$m.exit <- tillTid(lastseen) - tillTid(births$birthdate)
head(births[, c("id", "birthdate", "age", "exit", "m.exit", "lopnr", "antrec")])
summary(births[, c("age", "exit", "m.exit")])
with(births, sum(!is.na(m.exit) & (exit > m.exit)))
```

No surprise, errors in the raw data file; *UTDATMIN* is not calculated in a reliable way, since we have dates later than that with "activity". Let us try *UTDATMAX* instead:


```{r getpersonMAX}
lastseen <- person$UTDATMAX[match(births$id, person$ID)]
births$m.exit <- tillTid(lastseen) - tillTid(births$birthdate)
head(births[, c("id", "birthdate", "age", "exit", "m.exit", "lopnr", "antrec")])
summary(births[, c("age", "exit", "m.exit")])
with(births, sum(!is.na(m.exit) & (exit > m.exit)))
```

No big improvement. I follow the following rule for the *last* record for each woman: If *m.exit > 50*, set *exit = 50* (as it is);
if *is.missing(m.exit)* *exit == 50* (as it is); if *enter < m.exit < exit*, set *exit = m.exit*; if *m.exit < enter* set *m.exit = enter + 0.01*.

```{r setexit}
births$exit <- ifelse(with(births, !is.na(m.exit) & (m.exit > enter) & (lopnr == antrec)), 
                      pmin(births$exit, births$m.exit),
                      births$exit)
births$exit <- ifelse(with(births, !is.na(m.exit) & (m.exit < enter) & (lopnr == antrec)),
                      births$enter + 0.01,
                      births$exit)
head(births)
```

## Marital fertility

It is reasonable to concentate on *married at first birth only:

```{r, firstmar}
sel <- births$id[births$ab == "married" & (births$lopnr == 1)]
married <- births[births$id %in% sel, ]
married$birthdate <- tillTid(married$birthdate) - 1845
```


## A Glenn analysis

Have couples a preference for boys? One way to test for that is to see if the sex ratio affects the hazard of a new birth. I simlify somewhat and use as covariate *I(boys < girls)*. One should perhaps do this parity by parity, but by stratifying on parity we achieve that goal, with the constraint that we assume the effect to be independent of *parity*.

```{r stratage}
fit <- coxreg(Surv(enter, exit, event) ~ I(boys < girls) + sex + stillbirth + strata(parity) + birthdate, data = married)
summary(fit)
```

We can also do it by looking at interval length rather than age:

```{r ivll}
fit <- coxreg(Surv(exit - enter, event) ~ age + I(boys < girls) + sex + stillbirth + strata(parity) + birthdate, data = married)
summary(fit)
```

How should this be done?
